---
title: "Stat 133 HW04: String Manipulation and Regex"
author: "Your name and SID"
output: pdf_document
---

## Introduction

This assignment has two purposes:

a) to familiarize you with manipulating character strings
b) to introduce you to regular expressions in R

Submit your assignment to bcourses, specifically turn in your __Rmd__ (R markdown) file as well as the produced pdf file. Make sure to change the argument `eval=TRUE` inside every testing code chunk.


-----

## Names of Files

Imagine that you need to generate the names of 4 data files (with .csv extension). All the files have the same prefix name but each of them has a different number: `plot01.png`, `plot02.png`, `plot03.png`, and `plot04.png`. We can generate a character vector with these names in R. One naive solution would be to write something like this:
```{r}
files <- c('plot01.png', 'plot02.png', 'plot03.png', 'plot04.png')
```

Now imagine that you need to generate 100 file names. You could write a vector with 100 file names but it's going to take you a while.

How would you generate the corresponding character vector `files` in R containing 100 files names:  `plot01.png, plot02.png, plot03.png, ..., plot99.png, plot100.png`? Notice that the numbers of the first 9 files start with 0.

```{r}
# your code

```


-----

## USA States Names 

One of the datasets that come in R is `USArrests`. The row names of this data correspond to the 50 states. We can create a vector `states` with the row names:
```{r}
states <- rownames(USArrests)
head(states, n = 5)
```


### Using `grep()`

You can use the function `grep()` to know if there are any states containing the letter "z".
```{r}
# states containing the letter 'z'
grep(pattern = 'z', x = states)
```
In this case there is just one state (the third one) which corresponds to `r states[grep(pattern = 'z', x = states)]`

You can also use `grep()` with its argument `value = TRUE` to obtain the value of the matched pattern:
```{r}
# states containing the letter 'z'
grep(pattern = 'z', x = states, value = TRUE)
```

__Your turn__. Use `grep()`---and maybe other functions---to write the commands that answer the following questions:

- How many states contain the letter `i`?
- How many states contain the letter `q`?
- How many states do not contain the letter `a`?
- Which states contain the letter `j`?
- Which states contain the letter `x`?
- Which states are formed by two words?
- Which states start with `W` and end with a vowel?
- Which states start with `W` and end with a consonant?
- Which states contain at least three `i` (e.g. Illinois)?
- Which states contain five vowels (e.g. California)?
- Which states have three vowels next to each other (e.g. Hawaii)?

Tip: You can use `grep()`'s argument `ignore.case` to ignore letters in lower or upper case.
```{r}
# your answers
```

-----

## Starts with ... 

Write a function `starts_with()` such that, given a character string and a single character, it determines whether the string starts with the provided character. 

```{r}
# starts_with
```

Test it:
```{r, eval=FALSE}
starts_with("Hello", 'H')  # TRUE

starts_with("Good morning", 'H')  # FALSE
```

-----

## Ends with ... 

Now write a function `ends_with()` such that, given a character string and a single character, it determines whether the string ends with the provided character. 

```{r}
# ends_with
```

Test it:
```{r, eval=FALSE}
ends_with("Hello", 'o')  # TRUE

ends_with("Good morning", 'o')  # FALSE
```

-----

## Colors in Hexadecimal Notation

Write a function `is_hex()` that checks whether the input is a valid color in hexadecimal notation. Remember that a hex color starts with a hash `#` symbol followed by six hexadecimal digits: 0 to 9, and the first six letters A, B, C, D, E, F. Since R accepts hex-colors with lower case letters (a, b, c, d, e, f) your function should work with both upper and lower case letters.
```{r}
# is_hex()
```

Test it:
```{r, eval=FALSE}
is_hex("#ff0000")  # TRUE
is_hex("#123456")  # TRUE
is_hex("#12Fb56")  # TRUE

is_hex("#1234GF")  # FALSE
is_hex("#1234567") # FALSE
is_hex("blue")     # FALSE
```

-----

## Hexadecimal Colors with Transparency

Write a function `is_hex_alpha()` that determines whether the provided input is a hex color with alpha transparency. Remember that such a color has 8 hexadecimal digits instead of just 6.

```{r}
# is_hex_alpha()

```

Test it:
```{r, eval=FALSE}
is_hex_alpha("#FF000078")  # TRUE

is_hex_alpha("#FF0000")   # FALSE
```


## Hexadecimal Color split in RGB values

Write a function `hex_values()` that takes a hex-color and returns a vector with the values of the RGB---and possibly alpha---channels. 

Test it:

```{r, eval = FALSE}
# color with no transparency
hex_values("#435690")
```

If the provided color has an alpha channel, then the output should display such value:
```{r, eval = FALSE}
# color with transparency
hex_values("#435690FF")
```

If the input is not a valid hex-color, `hex_vales()` should return the message: `input is not a valid hexadecimal color`. The message should be displayed via `cat()`

```{r, eval = FALSE}
# invalid hex color
hex_values("#435XY90")
```

-----

## Splitting Characters

Create a function `split_chars()` that splits a character string into one single character elements.

```{r}
# split_chars()
```

Test it:
```{r, eval=FALSE}
split_chars('Go Bears!')

split_chars('Expecto Patronum')
```

Note that `split_chars()` returns the output in a single vector. Each element is a single character.

-----

## Number of Vowels

Create a function `num_vowels()` that returns the number of vowels of a character vector. In this case, the input is a vector in which each element is a single character.

```{r}
# num_vowels()
```

Test it:
```{r, eval=FALSE}
vec <- c('G', 'o', ' ', 'B', 'e', 'a', 'r', 's', '!')
num_vowels(vec)
```

Notice that the output is a numeric vector with five elements. Each element has the name of the corresponding vowel.

-----

## Counting Vowels

Use the functions `split_chars()` and `num_vowels()` to write a function `count_vowels()` that computes the number of vowels of a character string:

```{r}
# count_vowels()
```

Test it:
```{r, eval=FALSE}
count_vowels("The quick brown fox jumps over the lazy dog")
```

Make sure that `count_vowels()` counts vowels in both lower and upper case letters:

```{r, eval=FALSE}
count_vowels("THE QUICK BROWN FOX JUMPS OVER THE LAZY DOG")
```

-----

## Number of Consonants

Write a function `num_cons()` that counts the number of consonants regardless of whether thery are in upper or lower case (just the number, not the counts of each letter)
```{r, eval=FALSE}
# num_cons()
```

Test it:
```{r, eval=FALSE}
fox <- "The quick brown fox jumps over the lazy dog"
num_cons(fox)
```

-----

## Reversing Characters

Write a function that reverses a string by characters

```{r}
# reverse_chars()
```

Test it:
```{r, eval=FALSE}
reverse_chars("gattaca")
reverse_chars("Lumox Maxima")
```

-----

## Reversing Sentences by Words

Write a function `reverse_words()` that reverses a string (i.e. a sentence) by words

```{r}
# reverse_words()
```

Test it:
```{r, eval=FALSE}
reverse_words("sentence! this reverse")
```

If the string is just one word then there's basically no reversing:
```{r, eval=FALSE}
reverse_words("string")
```

-----

