---
title: "Lab 10"
author: "Your Name"
date: "July 16, 2015"
output: html_document
---

## Merging Data

### Appending Data

The easiest way to merge two datasets is to simply append one to the other, either horizontally or vertically.
Typically `cbind` or `rbind`, respectively, are used for this type of merge.

```{r}
A <- data.frame(city = c("Norfolk", "Indianapolis", "Boston", "Albuquerque"),
  state = c("Virginia", "Indiana", "Massachusetts", "New Mexico"))
B <- data.frame(city = c("Houston", "Atlanta", "Dallas", "Spokane"),
  state = c("Texas", "Georgia", "Texas", "Washington"))
horizAB <- cbind(A, B)
print(horizAB)
# In this case rbind probably makes more sense than cbind.
vertAB <- rbind(A, B)
print(vertAB)
```

Unfortunately this approach only works when both datasets have the same rows or columns.
It also gives undesirable results when there are repeated rows or columns.

```{r}
C <- data.frame(city = c("Norfolk", "Indianapolis", "Detroit", "Buffalo"),
  state = c("Virginia", "Indiana", "Michigan", "New York"))
D <- data.frame(city = c("Richmond", "Honolulu", "Norfolk", "Charleston"),
  state = c("Virginia", "Hawaii", "Virginia", "South Carolina"))
vertCD <- rbind(C, D)
# The first and seventh rows are identical.
print(vertCD)
```

In either of these cases, `merge` may be a better choice.

```{r}
# More on `all = TRUE"`later.
mergeCD <- merge(C, D, all = TRUE)
print(mergeCD)
```

### `merge`

The `merge` function combines two data frames by row.
By default, it keeps only those rows which match in corresponding columns.

```{r}
E <- data.frame(city = c("Norfolk", "Indianapolis", "Boston", "Albuquerque"),
  rain = c(44.7, 38.7, 42.5, 7.8))
# `merge` matches based on the columns with the same name, in this case `city`.
mergeAE <- merge(A, E)
print(mergeAE)
```

The `by` argument in `merge` allows you to specify which columns are matched.
If corresponding columns have different names in the two data frames, you can specify them separately using `by.x` (for the first data frame) and `by.y` (for the second).

Your turn!

The `precip` dataset in the base R package contains rainfall measurements for several US cities.
We want to merge part of with `A` and `B` to create a data frame containing city name, state name, and rainfall.

```{r, eval = FALSE}
# Remove `eval = FALSE` or set `eval = TRUE` when you have completed this code chunk.
# First examine the format of `precip`. You will find that it is a named vector.
str(precip) # Feel free to examine `precip` in other ways.
# In order to use `merge`, we need to create a data frame from `precip`.
precipdf <- # Fill this in.
# Now merge `precipdf` with data frames `A` and `B`.
# You should obtain a data frame with 8 rows and 3 columns.
# Fill this in.

```

The `all` argument in `merge` keeps all rows, even those which did not match; rows without a counterpart in the other data frame have NA for any additional columns.
Similar to `by`, `all.x` and `all.y` keep all rows from the first data frame and second data frame, respectively.

Repeat the previous merge, but keep all the cities in the `precip` dataset.

```{r}
# Fill this in.

```

### A More Complicated Task

Here are two datasets with some US Census data.

```{r}
pop77 <- state.x77
head(pop77)
library("readr")
pop14 <- read_csv("https://www.census.gov/popest/data/state/totals/2014/tables/NST-EST2014-01.csv", skip = 3)
head(pop14)
```

First create a data frame from `pop77`.
Make sure to include the state name as one of the columns.

```{r, eval = FALSE}
# Remove `eval = FALSE` or set `eval = TRUE` when you have completed this code chunk.
pop77df <- # Fill this in.
```

Now modify the `pop14` data frame so that it contains only the state names and the 2014 population data.

```{r}
# Fill this in.
```

Finally merge the two data frames into one containing all of the data from `pop77` in addition to the the 2014 population data.
The final result should contain 50 rows and 10 columns.

```{r}
# Fill this in.
```
