---
title: "strings"
date: "July 7, 2015"
output: html_document
---

Today we'll practice some basic string manipulations.

## Converting from Fahrenheit Degrees

Here are the four functions that we saw in HW3 that convert from Fahrenheit degrees to other temperature scales:

```{r}
to_celsius <- function(x) {
  (x - 32) * (5/9)
}

to_kelvin <- function(x) {
  (x + 459.67) * (5/9)
}

to_reaumur <- function(x) {
  (x - 32) * (4/9)
}

to_rankine <- function(x) {
  x + 459.67
}
```

We also saw how to use the previous functions to create a more general function ```convert()```:
```{r}
convert <- function(x, to = "celsius") {
  switch(to,
         "celsius" = to_celsius(x),
         "kelvin" = to_kelvin(x),
         "reaumur" = to_reaumur(x),
         "rankine" = to_rankine(x))
}

convert(30, 'celsius')
```

```convert()``` works fine when the argument ```to = 'celsius'```. What happens if you try ```convert(30, 'Celsius')``` or ```convert(30, 'CELSIUS')```?

__Your turn__. Rewrite ```convert()``` such that the argument ```to``` can be given in upper or lower case letters. For instance, the following three calls should be equivalent:

```{r, eval=FALSE}
convert(30, 'celsius')
convert(30, 'Celsius')
convert(30, 'CELSIUS')
```

```{r, eval=FALSE}
# function convert
convert <- function(x, to = "celsius") {
  # fill in
}

convert(30, 'celsius')
```

-----

## Names of files

Imagine that you need to generate the names of 10 data files (with .csv extension). All the files have the same prefix name but each of them has a different number: ```file1.csv```, ```file2.csv```, ... , ```file10.csv```. 

How can you generate a character vector with these names in R? Write a script with as many different ways to get such a vector:
```{r}
# vector of names
```

-----

## Using function ```cat()```

Run the following code:

```{r, eval=FALSE}
# name of output file
outfile <- "output.txt"

# writing to `outfile.txt```
cat("This is the first line", file = outfile)
# insert new line
cat("\n", file = outfile, append = TRUE)
cat("A 2nd line", file = "output.txt", append = TRUE)
# insert 2 new lines
cat("\n\n", file = outfile, append = TRUE)
cat("The quick brown fox jumps over the lazy dog",
    file = outfile, append = TRUE)
```

Look for the file ```output.txt``` in your working directory and open it. One of the uses of ```cat()``` is to write contents to a text file. Note that the first call to ```cat()``` does not include the argument ```append```. The rest of the calls do include ```append = TRUE```.

__Your turn__. Modify the script such that the content of ```output.txt``` looks like the header of an ```.Rmd``` file with your information:

```{r, eval= FALSE}
---
title: "Some title"
author: "Your name and SID"
date: "today's date"
output: pdf_document
---
```

-----

## Valid Color Names

Write a function ```is_color()``` to test if a given name---in English---is a valid R color. If the provided name is a valid R color, ```is_color()``` returns ```TRUE```. If the provided name is not a valid R color ```is_color()``` returns ```FALSE```.
```{r, eval=FALSE}
is_color <- function(x) {
  # fill in
  
}

# TRUE
is_color('yellow')

# FALSE
is_color('blu')

# FALSE
is_color('turkuiose')
```

-----

## Plot with a valid color

Use ```is_color()``` to create the function ```colplot()``` that takes one argument ```col``` (the name of a color) to produce a simple scatter plot. If ```col``` is a valid name---say "blue"---, the scatterplot should show a title ```"testing color blue"```. If the provided ```col``` is not a valid color name, e.g. "blu", then the function must stop, showing an error message ```"invalid color blu"```
```{r, eval=FALSE}
colplot <- function(col = "black") {
  # check that 'col' is valid
  
  # simple scatterplot
  plot(1:10, 1:10, col = col, pch = 19, cex = 2)
  # add title

}

# this should plot
colplot('tomato')

# this stops with error message
colplot('tomate')
```

-----

## Counting number of vowels

Consider the following vector ```lets``` which contains various letters:
```{r}
# vector of letters
set.seed(1)
lets <- sample(letters, size = 100, replace = TRUE)
head(lets)
```

If you were to count the number of vowels in ```lets``` you would get the following counts:

- a: 2
- e: 2
- i: 6
- o: 2
- u: 8

Write some instructions in R to count the number of vowels in vector ```lets```. 
```{r}
# count number of vowels
```

Test your script with ```lets``` and verify that you get the same counts for each vowel.


Now, take your script and adapt it into a function ```num_vowels()```

```{r}
# function num_vowels()
```

-----

## Number of letters, vowels, and consonants

Write a script that given a vector of letters (e.g. ```lets```) computes the total number of letters, the number of vowels, and the number of consonants. For instance, given the vector ```lets```, the script will print on console:

- ```"letters: 100"```
- ```"vowels: 20"```
- ```"consonants: 80"```

```{r}
# your script

```

