---
title: "Stat 133: Regex"
author: "Gaston Sanchez"
output: html_document
---

-----

## Hands-on with Regular Expressions

Packages for this practice:
```{r}
library(readr)
library(stringr)
```


## Bike Accidents in the Bay Area

Read the data `bike-accidents.csv` from the github repository:
```{r}
df <- read_csv("https://raw.githubusercontent.com/gastonstat/stat133/master/datasets/bike-accidents.csv")
df <- read.csv("~/Documents/stat133/stat133/datasets/bike-accidents.csv", stringsAsFactors = FALSE, nrows = 1000)

```

The first column in `bike-accidents.csv` is `DateTime` that contains the date and time of each event.
```{r}
head(df$DateTime)
```

__The ultimate goal is to separate dates from times (more on this later).__

-----

Some functions that you may need to use:

| Base R       | "stringr"       |
| -------------|-----------------|
| `paste()`    | `str_c()`       |
| `grep()`     | `str_extract()` |
| `substr()`   | `str_sub()`     |
| `sub()`      | `str_replace()` |
| `strsplit()` | `str_split()`   |
| `nchar()`    | `str_length()`  |


Regex quantifiers:

- `*`: matches at least 0 times.   
- `+`: matches at least 1 times.     
- `?`: matches at most 1 times.    
- `{n}`: matches exactly n times.    
- `{n,}`: matches at least n times.    
- `{n,m}`: matches between n and m times.     

-----

## Dates and Times

Let's start small by taking a sample of 10 values from the `DateTime` column:
```{r}
# sample of date-times
set.seed(10)
x <- sample(df$DateTime, size = 10)
x
```

Looking at the values in `DateTime` they all have a similar format. 


1. How would you get the __date__ values from `x`? e.g. `08/18/07`
```{r}
# dates from x (by position)
substr(x, start = 1, stop = 8)

# dates from x (regex pattern)
str_extract(x, '[0-9]{2}/[0-9]{2}/[0-9]{2}')
str_extract(x, '\\d{2}/\\d{2}/\\d{2}')
```

2. How would you get the __months__ from `x`? e.g. `08`
```{r}
# months from x
substr(x, start = 1, stop = 2)
```

3. How would you get the __days__ from `x`? e.g. `18`
```{r}
# days from x
substr(x, start = 4, stop = 5)
```

4. How would you get the __years__ from `x`? e.g. `07`
```{r}
# years from x
substr(x, start = 7, stop = 8)
```

5. How would you form __complete years__? e.g. `2007`
```{r}
# complete years
paste(20, substr(x, start = 7, stop = 8), sep = '')
```

6. How would you get the __times__ (hour and minutes): e.g. `7:55`?
```{r}
# times from x (by pattern)
str_extract(x, '[0-9]{1,}:[0-9]{2}')
str_extract(x, '[0-9]+:[0-9]{2}')

# by position
substr(x, 11, nchar(x) - 4)
```

7. How would you get the __time period__: e.g. `p.m.`?
```{r}
# time periods from x (by position)
str_sub(x, start = -4, end = -1)

# time periods from x (by pattern)
str_extract(x, '[ap].m.')
str_extract(x, '(a|p).m.')
```


-----

## Dates and Times

Now that you know how to get dates, times, and related values from `x`, let's work on the entire data frame and add new columns to `df`

1. Add a column `Date` with just the date values
```{r}
# add Date
df$Date <- substr(df$DateTime, start = 1, stop = 8)
```

2. Add a column `Month` with just the month number (as numeric)
```{r}
# add Month
df$Month <- as.numeric(substr(df$DateTime, start = 1, stop = 2))
```

3. Add a column `Day` with just the day number (as numeric)
```{r}
# add Day
df$Day <- as.numeric(substr(df$DateTime, start = 4, stop = 5))
```

4. Add a column `Year` with just the full year number (as numeric)
```{r}
# add Year
df$Year <- as.numeric(
  paste(20, substr(df$DateTime, start = 7, stop = 8), sep = '')
)
```

5. Add a column `Time` with just the time (NO time periods a.m. or p.m.)
```{r}
# add Time
df$Time <- str_extract(df$DateTime, '[0-9]+:[0-9]{2}')
```

6. Add a column `Period` with just the time period (a.m. or p.m.)
```{r}
# add Period
df$Period <- str_sub(df$DateTime, start = -4, end = -1)
```

7. Add a column `Hour` with just the hour in 24hr value (as numeric)
```{r}
# add Hour
df$Hour <- as.numeric(str_extract(df$Time, '[0-9]+'))
```

8. Add a column `Minute` with just the minute value (as numeric)
```{r}
# add Minute
df$Minute <- as.numeric(str_extract(df$Time, '[^:][0-9]+'))
```


-----

## Bonus questions

If you still have time left, here are some extra questions related to the type of road where accidents occurred (i.e. type of street):

- How many accidents took place on a __Boulevard__ in `Street1`?
```{r}
length(grep(' Bl', df$Street1))
```

- How many accidents took place on an __Avenue__ in `Street1` (e.g. Wilkinson Av)?
```{r}
length(grep(' Av', df$Street1))
```

- How many accidents took place on a __Drive__ in `Street1` (e.g. North Point Dr)?
```{r}
length(grep(' Dr', df$Street1))
```

- How many accidents took place on a __Road__ in `Street1` (e.g. South King Rd)?
```{r}
length(grep(' Rd', df$Street1))
```

- How many accidents took place on __numbered streets__ in `Street1` (e.g. 12th St, 13th Avenue) ?
```{r}
length(grep('[0-9]+', df$Street1))
```

