---
title: "parsing-xml"
author: "Gaston Sanchez"
date: "July 28, 2015"
output: html_document
---

-----

### Parsing XML content

Load the package "XML"

```{r}
library(XML)
```

Let's parse some xml content from [http://www.xmlfiles.com/examples/plant_catalog.xml](http://www.xmlfiles.com/examples/plant_catalog.xml). The default option is to use `xmlParse()`
```{r, eval=FALSE}
# xml file name
xml <- 'http://www.xmlfiles.com/examples/plant_catalog.xml'

# parse xml content (defult usage)
# (output as an internal C-structue)
doc1 <- xmlParse(xml)
```

Another alternative is to use `xmlTreeParse()` which returns an `"XMLDocument"` (i.e. an R hierarchical structure)
```{r, eval=FALSE}
# parse xml content
# (output as an R hierarchical structure)
doc2 <- xmlTreeParse(xml)
```

Once we have the two types of xml documents we can get access to the root node with `xmlRoot()`
```{r, eval=FALSE}
# access root node
top1 <- xmlRoot(doc1)
top2 <- xmlRoot(doc2)
```

Name of root node:
```{r, eval = FALSE}
xmlName(top1)
```

Names of nodes within root node:
```{r, eval = FALSE}
names(top1)
```

To get access to the children nodes of any node you can use `xmlChildren()`. Let's access the child nodes of the root:
```{r, eval=FALSE}
plants1 <- xmlChildren(top1)
plants2 <- xmlChildren(top2)
```

Let's fetch the data for the first child node (i.e. first plant)
```{r, eval=FALSE}
top1[[1]]   # top2[[1]]

# equivalently
plants1[[1]]
```

Likewise, we can use `xmlChildren()` to get the child nodes of the `top[[1]]`
```{r, eval=FALSE}
xmlChildren(top[[1]])
```

-----

### Inspecting a node

Let's get just the names of the nodes in the first plant
```{r eval=FALSE}
names(top1[[1]])   # names(top2[[1]])
```

Some functions to inspect nodes:
```{r, eval=FALSE}
# name of first child node
xmlName(top1[[1]])

# size of the first child node
xmlSize(top1[[1]])

# attributes of first child node
xmlAttrs(top1[[1]])   # there are no attributes
```

We can access the subchild nodes of the first plant:
```{r, eval=FALSE}
top1[[1]][["COMMON"]]


top1[[1]][["BOTANICAL"]]


top1[[1]][["ZONE"]]
```

-----

### Iterating on various nodes

How do we get the values "COMMON", "BOTANICAL", and "ZONE" all at once? One option is to use `lapply()` on `top2` (the object with a hierarchical R structure)
```{r, eval=FALSE}
lapply(top2[[1]], xmlValue)
```
`lapply()` does NOT work on `top1` because it is an internal C-structure (not an R structure)


To get the "COMMON" nodes of all plants we can use `apply()` on `plants2` like this:
```{r, eval=FALSE}
lapply(plants2, function(x) x['COMMON'])

# same for "PRICE" nodes
lapply(plants2, function(x) x['PRICE'])
```


Another option to get the same values is to use `xmlApply()` and `xmlSApply()` on both `top1` and `top2`
```{r, eval=FALSE}
# in list output
# same for: xmlApply(top2[[1]], xmlValue)
xmlApply(top1[[1]], xmlValue)

# in simplified lst output (usually a vector)
# same for: xmlSApply(top2[[1]], xmlValue)
xmlSApply(top1[[1]], xmlValue)
```

What happens if you use `xmlApply()` or `xmlSApply()` on the root node?
```{r, eval=FALSE}
# in list output
# same for: xmlApply(top2, xmlValue)
xmlApply(top1, xmlValue)

# in simplified lst output (usually a vector)
# same for: xmlSApply(top2, xmlValue)
xmlSApply(top1, xmlValue)
```

Loop over the child nodes (i.e. PLANTs) to get their sizes
```{r, eval=FALSE}
xmlSApply(top1, xmlSize)
```

Loop over the child nodes of the root to get the values:
```{r, eval=FALSE}
xmlSApply(top1, function(x) xmlSApply(x, xmlValue))
```

-----

A convenient function is `xmlToDataFrame()` which we can apply on  `top1`. `xmlToDataFrame()` can be used to extract data from an XML document (or sub-document) that has a simple, __shallow structure__.
```{r, eval=FALSE}
xmlToDataFrame(top1)
class(top1)
```
For more nested structures `xmlToDataFrame()` will probably not work as expected.

