---
title: "Stat 133: Regex"
author: "Gaston Sanchez"
output: html_document
---

## Getting Started with Regular Expressions

Here's a vector with some character strings:
```{r}
# vector of strings
animals <- c('dog', 'cat', 'bird', 'dolphin', 'lion',
             'zebra', 'tiger', 'wolf', 'whale', 'eagle',
             'pig', 'osprey', 'kangaroo', 'koala')
```


### Matching Literal Characters

Matching `dog`
```{r, eval=FALSE}
# 'dog'
grep('dog', animals, value = TRUE)
```

Animals containing letter `d`
```{r, eval=FALSE}
# 'd'
grep('d', animals, value = TRUE)
```


### Some Special Characters

The vertical bar is the __OR__ operator. Animals with letters `c` _OR_ `d`
```{r, eval=FALSE}
# 'c' or 'd'
grep('c|d', animals, value = TRUE)
```

Dot `.` represents __any__ character:
```{r, eval=FALSE}
# match any character
grep('.', animals, value = TRUE)
```


### Character Classes

Character classes are specified inside brackets `[]`. For example, here's how to get animals with letters `c` or `d`
```{r, eval=FALSE}
# 'c' or 'd'
grep('[cd]', animals, value = TRUE)
```

A caret `^` at the begining of a class indicates negation.
```{r, eval=FALSE}
# anything but 'cat'
grep('[^cat]', animals, value = TRUE)
```


### Position Matching

A caret `^` at the begining of a pattern is used to match characters in the starting position. To get animals starting with `d`
```{r, eval=FALSE}
# start with 'd'
grep('^d', animals, value = TRUE)
```

A `$` sign used at the end of the pattern indicates the last position. To get animals ending with `n`
```{r, eval=FALSE}
# end with 'd'
grep('n$', animals, value = TRUE)
```


### Quantifiers

Quantifiers specify how many repetitions of the pattern.   

  * `*`: matches at least 0 times.   
  * `+`: matches at least 1 times.     
  * `?`: matches at most 1 times.    
  * `{n}`: matches exactly n times.    
  * `{n,}`: matches at least n times.    
  * `{n,m}`: matches between n and m times.     

```{r, eval=FALSE}
# at least 0 or more 'o'
grep('o*', animals, value = TRUE)

# at least 1 'o'
grep('o+', animals, value = TRUE)

# at most 1 'o'
# (what's hapenning here?)
grep('o?', animals, value = TRUE)

# exactly 2 'o'
grep('o{2}', animals, value = TRUE)
```


### More examples

Animals with one `o`, but not two `o`'s together
```{r, eval=FALSE}
# animals with no two 'o' together
grep('o.[^o]', animals, value = TRUE)
```

Animals containing both ```a``` and ```e```
```{r, eval=FALSE}
# containing both 'a' and 'e'
grep('(a.*e)|(e.*a)', animals, value = TRUE)
```

Animals with more than 1 vowel
```{r, eval=FALSE}
grep('[aeiou].*[aeiou]', animals, value = TRUE)
```

If you want to get animals that don't contain `d`:
```{r, eval=FALSE}
# which names have 'd'
grep('d', animals)
# excluding those names with 'd'
animals[-grep('d', animals)]
```



-----

## More Regex

Here's another vector with character strings
```{r}
files <- c('sales1.csv', 'orders.csv', 'sales2.csv',
           'sales3.csv', 'europe.csv', 
           'na1.csv', 'na2.csv', 'sa.csv', 'ca.csv')
```


### Character and POSIX classes

```{r, eval=FALSE}
# file names with numbers
grep('[0123456789]', files, value = TRUE)

# equivalently
grep('[0-9]', files, value = TRUE)

# equivalently
grep('[[:digit:]]', files, value = TRUE)
```


### Replacing characters

```{r, eval=FALSE}
# changing 'csv' extension into 'txt' extension
sub('csv', 'txt', files)
```

```{r, eval=FALSE}
# separate file name from file extension
# (we need to escape the dot character)
strsplit(files, '\\.')
```


