---
title: "Stat 133 HW03: Flow Control Structures and Functions with R"
author: "Your name and SID"
output: pdf_document
---

## Introduction

This assignment has two purposes:

a) to familiarize you with control flow structures in R 
b) to introduce you to writing functions in R

Submit your assignment to bcourses, specifically turn in your __Rmd__ (R markdown) file as well as the produced pdf file. Make sure to change the argument ```eval=TRUE``` inside every testing code chunk.

-----

## Last Element

Write a function ```last()``` that takes a vector (or factor) and returns the last element in the vector. For instance:
```{r}
# write your function
# last()

```

Test it:
```{r, eval = FALSE}
last(c('A', 'E', 'I', 'O', 'U'))
last(c(2, 4, 6, 8, 10))
```


-----

## If-then-else

Write a function ```multfive()``` that takes a number and determines whether the number is multiple of 5. If the provided number is multiple of five, then the output must be: ```it is multiple of five```. Conversely, if the provided number is not a multiple of five, then the output must be: ```it's not a multiple of 5```. For example:
```{r}
# write your function
# multfive()

```

Test it:
```{r, eval=FALSE}
# multiple of five
multfive(10)

# not a multiple of five
multfive(33)
```


-----

## Create your histogram plotting function

Write a function ```histogram()``` that plots a histogram with added vertical lines for the following summary statistics: minimum value, median, mean, and maximum value. The main idea is to wrap the high-level function ```hist()``` and then plot the lines with a low-level plotting function.

Define your function with the following requirements: 

- bars of histogram colored in "gray90""
- line of minimum value in color "gray30", and dashed type
- line of maximum value in color "gray30", and dashed type
- line of median value in color "orange"
- line of mean value in color "tomato"
- all lines (min, max, median, mean) with a width of 3
```{r}
# write your function
# histogram

```

Test it:
```{r, fig.width=5, fig.height=4, eval = FALSE}
histogram(mtcars$mpg)
```


-----

## Converting Fahrenheit Degrees

The table below shows the different formulas for converting Fahrenheit degrees into other scales:

| Units   | from Fahrenheit     |
| ------- | ------------------- |
| Celsius | (째F - 32) x 5/9     |
| Kelvin  | (째F + 459.67) x 5/9 |
| Reaumur | (째F - 32) x 4/9     |
| Rankine | 째F + 459.67         |

Write a function that converts from Fahrenheit degrees into each type of the four alternative scales. This implies writing four different functions:

- ```to_celsius()```
- ```to_kelvin()```
- ```to_reaumur()```
- ```to_rankine()```

```{r}
# write your functions

```

Test them:
```{r, eval=FALSE}
to_celsius(34)
to_kelvin(34)
to_reaumur(34)
to_rankine(34)
```

-----

## Using switch()

Create a function ```convert()``` that converts Fahrenheit degrees into the specified scale. Use ```switch()``` and the previously defined functions---```to_celsius()```, ```to_kelvin()```, ```to_reaumur()``` and ```to_rankine()```---to define ```convert()```. Use two arguments: ```x``` and ```to```, like this:
```{r}
# write your function
# convert()

```

Test it:
```{r, eval=FALSE}
convert(32, "celsius")
convert(32, "kelvin")
convert(32, "reaumur")
convert(32, "rankine")
```


-----

## Permutations

Write a function ```permute()``` that calculates the number of permutations of $k$ objects from a set of $n$ objects. 
```{r}
# write your function
# permute()

```

Test it:
```{r, eval=FALSE}
permute(6, 2)
```

Make sure that the function checks that both $n$ are $k$ are non-negative numbers (if any of them is negative, the function must stop). Also make sure that if $n$ is less than $k$, the result is zero. In addition, $n$ and $k$ should be coerced as integers.
```{r, eval=FALSE}
# the following calls should not work
permute(2, 6)
permute(-6, 6)
```


-----

## Average function with for loop

Create a function ```average()``` using a ___for loop___ to compute the mean. ```average()``` takes a numeric vector and returns the average.
$$
\bar{x} = \frac{1}{n} \sum^{n}_{i = 1} x_i
$$
```{r}
# write your function
# average()

```

Test it:
```{r, eval=FALSE}
average(1:5)
mean(1:5)
```


-----

## Geometric Mean function

The formula of the geometric mean is:

$$
\left( \prod^{n}_{i = 1} x_i \right)^{1/n}
$$

Write a function ```geomean()``` that computes the geometric mean of a vector of positive numbers, using a ___for loop___:
```{r}
# write your function
# geomean()

```

Test it:
```{r, eval=FALSE}
geomean(1)
geomean(1:5)
```


-----

## Frequency Table

Write a function ```freq_table()``` that takes a factor and generates a frequency table with 5 columns: 

1) ```category```: the levels of the factor
2) ```count```: absolute frequency
3) ```prop```: relative frequency (use four decimal places)
4) ```cumcount```: cumulative absolute frequency
5) ```cumprop```: cumulative relative frequency (use four decimal places)

Make sure that the input is a factor (otherwise the function should stop). Likewise, the output should be in ```data.frame``` form.

```{r}
# write your function
# freq_table()

```

Test it:
```{r, eval=FALSE}
# some factor
set.seed(13)
sizes <- factor(
  sample(c('small', 'medium', 'large'), size = 90, replace = TRUE)
)

# frequency table
freq_table(sizes)
```


-----

## Summary Statistics Table

Write a function ```stats()``` that takes a numeric vector and generates the following descriptive statistics: 

- ```min```: minimum value
- ```max```: maximum value
- ```range```: range (max - min)
- ```q1```: first quartile
- ```q3```: third quartile
- ```iqr```: inter-quartile range (q3 - q1)
- ```median```: median
- ```mean```: mean
- ```sd```: standard deviation
- ```NAs```: number of missing values ```NA```

The function ```stats()``` should include an argument ```na.rm``` ---that takes a logical value--- so it can handle potential missing values. The output must be a data.frame of one column.

```{r}
# write your function
# stats()

```

Test it:
```{r, eval=FALSE}
# no missing values
stats(1:10)
# missing values
stats(c(1:4, NA, 6:9, NA), na.rm = TRUE)
```


-----

## Frequency Table and Summary Statistics

Having created the functions ```freq_table()``` and ```stats()```, use them to write a function ```univarite()``` for producing summary statistics depending on the type of input. If the provided input is a numeric vector, then ```stats()``` should be called. In turn, if the provided input is a factor, then ```freq_table()``` should be called. If the input is not a numeric vector or a factor, then ```univariate()``` will print: ```"x must be either a numeric vector or a factor"```

```{r}
# write your function
# univariate()
```

Test it:
```{r, eval=FALSE}
# factor input
univariate(sizes)

# numeric input
univariate(1:10)
```

This should not work:
```{r, eval=FALSE}
# this should cause an error
univariate(colors()[1:5])
```


