---
title: "Stat 133, HW3"
author: "Your Name"
output: pdf_document
---

## Introduction

This assignment has two purposes:

a) to familiarize you with control flow structures in R 
b) to introduce you to writing functions in R

Submit your assignment to bcourses, specifically turn in your __Rmd__ (R markdown) file as well as the produced pdf file. Make sure to change the argument `eval=TRUE` inside every code chunk.


-----

## Arithmetic average with loops

R provides the function `mean()` that can be applied to calculate the arithmetic mean (i.e. average) of a numeric vector. Here's the formula of the average for a vector $x = (x_1, x_2, \dots, x_n)$
$$
\bar{x} = \frac{1}{n} \sum^{n}_{i = 1} x_i
$$

Use the three types of loops---`for, repeat, while`---to get the average of the numeric vector  `x <- 1:10`

```{r}
# your for loop




# your repeat loop




# your while loop




#
```


-----

## Area of a circle

For a given circle of radius $r$, the area $A$ is:
$$
A = \pi r^2
$$

Write a function `circle_area()` that calculates the area of a circle. This function must take one argument `radius`. Give `radius` a default value of 1.
```{r}
# write your circle_area() function


# 
```

Test it:
```{r, eval = FALSE}
# default (radius 1)
circle_area()

# radius 3
circle_area(radius = 3)
```


## Area of a cylinder

For a given cylinder of radius $r$ and height $h$ the area $A$ is:
$$
A = 2 \pi r h + 2 \pi r^2
$$

Notice that the formula of the area of a cylinder includes the area of a circle: $\pi r^2$. Write a function `cyl_area()`, that calls `circle_area()`, to compute the area of a cylinder. This function must take two arguments: `radius` and `height`. Give both arguments a default value of 1.
```{r}
# your cyl_area() function




# 
```

Test it:
```{r, eval = FALSE}
# default (radius 1, height 1)
cyl_area()

# radius 2, height 3
cyl_area(radius = 2, height = 3)
```


## Volume of a cylinder

For a given cylinder of radius $r$ and height $h$ the volume $V$ is:
$$
V = \pi r^2 h
$$

Write a function `cyl_volume()`, that calls `circle_area()`, to compute the volume of a cylinder. This function must take two arguments: `radius` and `height`. Give both arguments a default value of 1.

```{r, echo=FALSE}
# your cyl_volume() function



#
```

Test it:
```{r, eval = FALSE}
# default (radius 1, height 1)
cyl_volume()

cyl_volume(radius = 3, height = 10)
cyl_volume(height = 10, radius = 3)
```


-----

## Even number

Write a function `is_even()` that determines whether a number is even (i.e. multiple of 2). If the inpute number is even, the output should be `TRUE`. If the input number is odd, the output should be `FALSE`. If the input is not a number, the output should be `NA` 
```{r}
# your is_even() function




#
```

Test it:
```{r, eval = FALSE}
# even number
is_even(10)

# odd number
is_even(33)

# not a number
is_even('a')
```


## Odd number

Use your function `is_even()` to write a function `is_odd()` that determines if a number is odd (i.e. not a multiple of 2). If a number is odd, the output should be `TRUE`; if a number is even the output should be `FALSE`; if the input is not a number the output should be `NA`
```{r}
# your is_odd() function



#
```

Test it:
```{r, eval = FALSE}
# odd number
is_odd(1)

# even number
is_odd(4)

# not a number
is_odd('a')
```


-----

## Quadratic Formula

One way to find the real roots of a 2nd degree polynomial, $ax^2 + bx + c$, is by using the famous quadratic equation:
$$
x = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}
$$

Write a function `quadratic()` that solves the quadratic equation. It should take three arguments: `a, b, c` corresponding to the coefficients of the polynomial. Give default values `a=1, b=1, c=0`. The output must be a list with two elements `$sol1` and `$sol2` containing the two roots:
```{r}
# your quadratic() function



#
```

Test it:
```{r, eval = FALSE}
# default
quadratic(a = 1, b = 1, c = 0)

# x^2 + 2b - 2
quadratic(c = -2, b = 2, a = 1)

# complex roots
quadratic(a = 1, b = 2, c = 2)
```

In order to limit the results to just real roots, it would be nice to have a function that indicates that there are no real roots. Use your `quadratic()` function to write a function `find_roots()` that either: a) finds the real roots, or b) returns a message `"No real roots"` if there are no real roots
```{r}
# your find_roots() function




#
```

Test it:
```{r, eval = FALSE}
find_roots(a = 1, b = -1, c = -2)
find_roots(a = 1, b = 2, c = 2)
```


-----

## Converting Miles to other units

The table below shows the different formulas for converting miles  into other scales:

| Units   | Formula         |
| ------- | --------------- |
| Inches  | mi x 63360      |
| Feet    | mi x 5280       |
| Yards   | mi x 1760       |
| Meters  | mi / 0.00062137 |
| Kms     | mi / 0.62137    |

Write the following five functions for each type of conversion. Each function must take one argument `x` with default value `x = 1`:

- `miles2inches()`
- `miles2feet()`
- `miles2yards()`
- `miles2meters()`
- `miles2kms()`

```{r}
# write your functions





#
```

Test them: 
```{r, eval = FALSE}
miles2inches(2)
miles2feet(2)
miles2yards(2)
miles2meters(2)
miles2kms(2)
```

-----

## Using switch()

Create a function `convert()` that converts miles into the specified units. Use `switch()` and the previously defined functions---`miles2inches()`, miles2feet(), ..., `miles2kms`---to define `convert()`. Use two arguments: `x` and `to`, like this:
```{r, eval=FALSE}
convert(40, to = "in")
```

By default, `to = "km"`, but it can take values such as `"in"`, `"ft"`, `"yd"`, or `"m"`.
```{r}
# your convert() function



#
```

Test it:
```{r eval=FALSE}
convert(3, "in")
convert(3, "ft")
convert(3, "yd")
convert(3, "m")
convert(3, "km")
```


-----

## Create your histogram plotting function

Write a function ```histogram()``` that plots a histogram with added vertical lines for the following summary statistics: minimum value, median, mean, and maximum value. The main idea is to wrap the high-level function ```hist()``` and then plot the lines with a low-level plotting function.

Define your function with the following requirements: 

- bars of histogram colored in `"#DDDDDD"`
- line of minimum value in color `"#333333"`, and dotted type
- line of maximum value in color `"#333333"`, and dotted type
- line of median value in color `"#00CCFF"`
- line of mean value in color `"#FF7502"`
- max and min lines with a width of 2
- median and mean lines with a width of 3
- axis tick labels in horizontal orientation

```{r}
# your histogram() function





#
```

Test it:
```{r, eval = FALSE, fig.width=5, fig.height=4}
histogram(mtcars$hp)
```


-----

## Frequency Table

Write a function `freq_table()` that takes a factor and generates a frequency table with 5 columns: 

1) `category`: the levels of the factor
2) `frequency`: absolute frequency
3) `proportion`: relative frequency (round to four decimal digits)
4) `cumfreq`: cumulative absolute frequency
5) `cumprop`: cumulative relative frequency (round to four decimal digits)

`freq_table()` must have two arguments: `x` and `decreasing`. The argument `x` is the provided factor. The argument `decreasing` indicates whether to return the results of the table in decreasing or increasing order. Give `decreasing` a default value of `NULL`.

By default `decreasing = NULL` means that the table is returned as is. `decreasing = TRUE` indicates that the results are displayed by frequency in decreasing order; `decreasing = FALSE` indicates that the results are displayed by frequency in increasing order.
Make sure that the input is a factor (otherwise the function should convert the input into factor). Likewise, the output should be in `data.frame` format.


```{r}
# your freq_table() function




#
```

Test it:
```{r, eval = FALSE}
# some factor
set.seed(13)
sizes <- factor(
  sample(c('small', 'medium', 'large'), size = 90, replace = TRUE)
)

# frequency table (default)
freq_table(sizes)

# frequencies in decreasing order
freq_table(sizes, decreasing = TRUE)

# frequencies in increasing order
freq_table(sizes, decreasing = FALSE)
```

