---
title: "Lab 11"
author: "Your Name"
date: "July 23, 2015"
output: html_document
---

## Unix and Command Line

Today's lab should be shorter than normal.
We'll simply be trying to build a little familiarity with basic UNIX commands.
We'll also include analogous R commands where possible.

Before we begin, first open Terminal (if using Mac) or something similar---like Git Bash---(if using Windows).
Navigate (using `cd`) to the directory containing this file.
Also set your current working directory in RStudio to the one containing this file (using `setwd()`).
If you are using a relative path, `pwd` (in Terminal, etc.) or `getwd()` (in RStudio) can be used to find your current directory.

```{r, eval = FALSE}
getwd()
setwd(dir) # Replace dir with a path.
```

Let's create an empty text file in Terminal, etc.
Call it `test.txt`.
The `touch` command simply updates the access or modification time of a directory or file.
If the file does not exist, it creates it.

In R, the following command can be used to do the same.

```{r, eval = FALSE}
file.create("test.txt")
```

We can also create directories.
The command `mkdir` creates a new directory.
Create a subdirectory called `lab11`.

In R, we can do the same.

```{r, eval = FALSE}
dir.create("lab11")
```

Now let's move some stuff around.
In Terminal, etc., the commands `mv`, `cp`, and `rm` can be used.
For example, The command `mv test.txt test2.txt` renames `test.txt` to `test2.txt`.
The command `cp test2.txt test.txt` makes a copy of `test2.txt` and calls it `test.txt`.
In both cases, if the second file already exists, it is overwritten.
The command `rm test2.txt` would delete `test2.txt`; the command `rm -r lab11` would delete the subdirectory `lab11` and all of its contents.

Analagous commands in R also exist:

```{r, eval = FALSE}
file.rename("test.txt", "test2.txt")
file.copy("test2.txt", "test.txt")
file.remove("test2.txt")
unlink("lab11", recursive = TRUE)
```

We can also modify text files, without using a text editor.

Using `>` or `>>` will take the output of a command and place it in a given text file.
Use `>>` to append text, and `>` to overwrite any existing text.

For example, first type the command `echo Hello world!`.
Now create a text file named `test2.txt` containing the text `mkdir lab11`.

In R:

```{r, eval = FALSE}
cat('mkdir lab11', file = "test2.txt")
# To append text instead of overwriting it, set append = TRUE.
```

You should notice that the file just created actually contains a UNIX command.
If you call `test2.txt` in Terminal, etc., its contents will be run.
In this case the subdirectory `lab11` will be created.

Your turn!

Modify `test.txt` so that, when run, it performs the following actions:
  1.  Makes a new subdirectory called `testdir`.
  2.  Copies itself into that directory.
  3.  Changes the current directory to the new subdirectory.
  4.  Prints `Done!` when finished.
Then run it once to make sure it works.

For fun: Create a file `test3.txt` that prints `Looping ...` in an infinite loop.
