---
title: "Lab 11"
author: "Your Name"
date: "November 18, 2015"
output: html_document
---

## Unix and Command Line

We'll simply be trying to build a little familiarity with basic UNIX commands.
We'll also include analogous R commands where possible.

Before we begin, first open Terminal (if using Mac) or something similar---like Git Bash---(if using Windows).
Navigate (using `cd`) to the directory containing this file.
Also set your current working directory in RStudio to the one containing this file (using `setwd()`).
If you are using a relative path, `pwd` (in Terminal, etc.) or `getwd()` (in RStudio) can be used to find your current directory.

```{r, eval = FALSE}
getwd()
setwd(dir) # Replace dir with a path.
```

Let's create an empty text file in Terminal, etc.
Call it `test.txt`.
The `touch` command simply updates the access or modification time of a directory or file.
If the file does not exist, it creates it.

In R, the following command can be used to do the same.

```{r, eval = FALSE}
file.create("test.txt")
```

We can also create directories.
The command `mkdir` creates a new directory.
Create a subdirectory called `lab11`.

In R, we can do the same.

```{r, eval = FALSE}
dir.create("lab11")
```

Now let's move some stuff around.
In Terminal, etc., the commands `mv`, `cp`, and `rm` can be used.
For example, The command `mv test.txt test2.txt` renames `test.txt` to `test2.txt`.
The command `cp test2.txt test.txt` makes a copy of `test2.txt` and calls it `test.txt`.
In both cases, if the second file already exists, it is overwritten.
The command `rm test2.txt` would delete `test2.txt`; the command `rm -r lab11` would delete the subdirectory `lab11` and all of its contents.

Analagous commands in R also exist:

```{r, eval = FALSE}
file.rename("test.txt", "test2.txt")
file.copy("test2.txt", "test.txt")
file.remove("test2.txt")
unlink("lab11", recursive = TRUE)
```

We can also modify text files, without using a text editor.

Using `>` or `>>` will take the output of a command and place it in a given text file.
Use `>>` to append text, and `>` to overwrite any existing text.

For example, first type the command `echo Hello world!`.
Now create a text file named `test2.txt` containing the text `mkdir lab11`.

In R:

```{r, eval = FALSE}
cat('mkdir lab11', file = "test2.txt")
# To append text instead of overwriting it, set append = TRUE.
```

You should notice that the file just created actually contains a UNIX command.
If you call `test2.txt` in Terminal, etc., its contents will be run.
In this case the subdirectory `lab11` will be created.

Your turn!

Modify `test2.txt` so that, when run, it performs the following actions:
  1.  Makes a new subdirectory called `testdir`.
  2.  Copies itself into that directory.
  3.  Changes the current directory to the new subdirectory.
  4.  Prints `Done!` when finished.
Then run it once to make sure it works.


Practicing UNIX commands:

Create a new directory:
```{r eval=FALSE}
mkdir practice
```

Change to the created directory
```{r eval=FALSE}
cd practice
```

Let's download the dataset `starwars.csv` from the github repo. Use `curl` followed by the URL of the data file and indicate an output file (e.g. `stars.csv`): 
For more details, please check http://curl.haxx.se/docs/manpage.html
```{r eval=FALSE}
curl "https://raw.githubusercontent.com/gastonstat/stat133/master/datasets/starwars.csv" -o stars.csv
```

Now check the contents of `practice`
```{r eval=FALSE}
ls 
```

You can also use `curl` to download images:
```{r eval=FALSE}
curl 'http://40.media.tumblr.com/tumblr_m4ihtxcV021rprzhqo1_500.jpg' -o vader.jpg
```

Use `open` to see the downloaded image
```{r eval=FALSE}
open vader.jpg
```

-----

## Basic inspection

Let's start inspecting the file `stars.csv`.

`cat` allows you to display the contents of the file on the terminal:
```{r eval=FALSE}
cat stars.csv
```

The commands `head` and `tail` let you look at the first and last observations:
```{r eval=FALSE}
# first 5 rows
head -5 stars.csv

# last 4 rows
tail -4 stars.csv
```

`wc` (word count) gives you information about the number of characters, lines, and words
```{r eval=FALSE}
# default
wc stars.csv

# number of characters
wc -c stars.csv

# number of lines (i.e. rows)
wc -l stars.csv

# number of words (not very useful in this case)
wc -w stars.csv
```

----- 

## Using `grep`

Search for different individuals:
```{r eval=FALSE}
# female subjects
grep female stars.csv

# droids
grep droid stars.csv
```

__Your turn__: Get the rows for:
```{r, eval=FALSE}
# jedis

# male subjects

```

----- 

## Using `cut`

`cut` allows you to retrive specific columns from a text file with delimiters:

| Options | Description                                 |
| ------- | ------------------------------------------- |
| -f 1,3  | Return columns 1 and 3, delimited by tabs   |
| -d ","  | use commas as delimiters instead of tabs    |
| -c 3-8  | return characters 3 through 8 from the file |

For informaiton about cut: http://ss64.com/bash/cut.html
To get the first column type:
```{r, eval=FALSE}
# first column
cut -f 1 -d ',' stars.csv
```

Columns 1 and 4:
```{r, eval=FALSE}
# name and planet
cut -f 1,4 -d ',' stars.csv
```

----- 

## File Redirection

Let's subset the dataset and create a new file for non-humans
```{r, eval = FALSE}
grep human stars.csv > humans.csv
```

__Your turn__: create a new file `blaster.csv` containing those individuals with weapon "blaster" 
```{r, eval = FALSE}
# your command

```

----- 

## Pipe Redirection

Pipes allow us to combine different commands
```{r, eval = FALSE}
# names of first 5 rows
head -5 stars.csv | cut -f 1 -d ","
```

If we want to exclude the name of the column `name` we can do something like this:
```{r, eval = FALSE}
# names of first 5 rows
head -6 stars.csv | tail -5 | cut -f 1 -d ","
```

Get the name of female subjects 
```{r, eval = FALSE}
# names of female subjects
grep female stars.csv | cut -f 1 -d ","
```


----- 

## Filter `sort`

Sorting lines with `sort` 

| Options | Description                                           |
| ------- | ----------------------------------------------------- |
| -n      | Sort by numeric value rather than alphabetically      |
| -r      | Sort in inverse order                                 |
| -k 3    | Sort lines based on column 3 (space or tab delimiter) |
| -t ","  | Use commas for delimiters instead of tabs or spaces   |
| -u      | Return only a single unique representative            |

```{r, eval = FALSE}
# sorting names
cut -f 1 -d "," stars.csv | sort

# sorting names in reverse order
cut -f 1 -d "," stars.csv | sort -r
```

__Your turn__: Select name and height, and sort lines by height
```{r, eval = FALSE}
# your command


```

----- 

## Filter `uniq`

Isolating unique lines with `uniq`: Another useful command for extracting a subset of values from a file is `uniq`. The way `uniq` works is by removing __consecutive__ identical lines from a file, leaving one unique representative.

| Options | Description                                           |
| ------- | ----------------------------------------------------- |
| -c      | Count the number of occurrences of each unique line   |
| -f 4    | Ignore the first 4 fields in determinind uniqueness   |
| -i      | Ignore case when determining uniqueness               |

One attempt to get unique species would be:
```{r, eval = FALSE}
# unique species (column 12)
cut -f 12 -d "," stars.csv | uniq
```
But it doesn't work since values not all values appear consecutively

So first we need to `sort` them:
```{r, eval = FALSE}
# unique species
cut -f 12 -d "," stars.csv | sort | uniq
```

We can even get some frquencies:
```{r, eval = FALSE}
# counts of species
cut -f 12 -d "," stars.csv | sort | uniq -c
```

Notice that in the obtained frequencies there's the label `species`. To get rid of it we can add another pipe:
```{r, eval = FALSE}
# counts of species
tail -20 stars.csv | cut -f 12 -d "," | sort | uniq -c
```

-----

## Test yourself

```{r, eval = FALSE}
# how many humans

# how many jedis

# names of jedis

# names of jedis in alphabetical order

# table of frequencies for weapons
 

# table of frequencies for weapons in decreasing order

# ---
```
