---
title: "Lab 7"
author: "Your Name"
date: "October 13th, 2015"
output: html_document
---

## Regular Expressions

### Preliminaries

First, load the data.

```{r}
library("readr")

flags <- read_csv("https://raw.githubusercontent.com/gastonstat/stat133/master/datasets/flags.csv")
```

We'll be working only with the first column (the `name` column), which contains the names of countries.

```{r}
countries <- flags$name

head(countries)
```

You may have noticed that in the `countries` vector, words in a country's name are separated by `-`.
Let's replace them with spaces, so that for example `American-Samoa` becomes `American Samoa`. (More on this later.)

```{r}
# The syntax is gsub(pattern, replacement, x).
countries <- gsub("-", " ", countries)

# We also could have done chartr("-", " ", countries).

head(countries)
```

### grep

grep is the most basic regex function: it simply returns which elements of a vector match a particular pattern.
For example, suppose we wanted to know which countries have the word `islands` in their name.

```{r}
grep("islands", countries)
```

It would appear that there are none. But that's not actually true.

```{r}
countries[33]
```

So grep failed to match the pattern `islands` to the word `Islands`. There are a couple of ways to fix that.

```{r}
grep("Islands", countries)

# Or
grep("islands", countries, ignore.case = TRUE)
```

The second way has the advantage of being able to match both `islands` and `Islands`, if necessary, as well as `IsLands`, etc.

You can also view the country names instead of their indices.

```{r}
grep("islands", countries, ignore.case = TRUE, value = TRUE)
```

Your turn!
Use grep to find which countries have more than one word in their name. (Hint: What character is present in that case?)

```{r}
# Fill in.
```

### grep Variants

There are several variants of grep suitable for other types of string manipulation.

grepl is identical to grep, but returns a corresponding logical vector instead of a numeric vector.

regexpr and gregexpr are useful for finding the particular location of a match.
regexpr gives the index of the first match (if any) in each element of a vector, while gregexpr gives the indices of all matches and returns a list.

```{r}
regexpr("an", head(countries), ignore.case = TRUE)

gregexpr("an", head(countries), ignore.case = TRUE)
```

sub and gsub are useful for replacing matches.
sub replaces the first match (if any) in each element of a vector, while gsub replaces all matches.

```{r}
sub("an", "BANANA", head(countries), ignore.case = TRUE)

gsub("an", "BANANA", head(countries), ignore.case = TRUE)
```

Your turn!
Replace all instances of `us` (or `US`, etc.) with `United States`, but only output the countries whose names were changed.
(Hint: How can you find which countries contain `us`?)

```{r}
# Fill in.
```

### Metacharacters

So far we only to know how to match literal strings, except for ignoring case.
What if we wanted to know which countries' names start with `A`?
The metacharacter `^` denotes the beginning of the string.

```{r}
grep("^A", countries, value = TRUE)
```

Similarly, the metacharacter `$` denotes the end of the string.

Your turn!
Find which countries' names end in `stan`.

```{r}
# Fill in.
```

You may have noticed that metacharacters are not treated literally.
If you want to match a metacharacter, make sure to precede it by `\\`, as in `\\^'.
If you want to match `\`, use `\\\\`.

```{r}
# Switch dollar sign to letter s:
word <- "Bo$$"
gsub("$", "s", word) #wrong
gsub("\\$", "s", word) #correct
```

Suppose you wanted to know which countries' names start with `A` and whose third letter is `g`, but you didn't care about the second letter? The metacharacter `.` represents any character.

```{r}
grep("^A.g", countries, value = TRUE)
```

The metacharacters `?`, `*`, and `+` modify the previous item: `?` matches zero or one times, `*` matches any number of times (including zero), and `+` matches at least once.
So `.*` matches any character string, while `.+` matches any non-empty character string.

Your turn!
Find which countries' names start with `S` and have an `l` in their name.

```{r}
# Fill in.
```

### Character Classes

What if you wanted to know which countries' names start with `A` or `Z`? The metacharacter `|` functions as a logical or.
Left and right parentheses (`(` and `)`) are also metacharacters used for grouping.

```{r}
grep("^(A|Z)", countries, value = TRUE)
```

If you want to do this for, say, any vowel, you would have to write `(A|E|I|O|U)`, for example.
Character classes simplify this: the character class for the preceding expression is just `[AEIOU]`.
The previous code chunk can be equivalently written as follows:

```{r}
grep("^[AZ]", countries, value = TRUE)
```

Several character classes are predefined in R.
For example, `[:digits:]` is equivalent to `0123456789`, and `[:lower:]` is equivalent to `abcdefghijklmnopqrstuvwxyz`.
So the character class `[0123456789]` can be more easily written as `[[:digits:]]`, etc.
See the regex documentation for R for a more extensive list of predefined character classes.

Your turn!
Find which countries' names start with two or more vowels.

```{r}
# Fill in.
```

Finally, when inside of a character class, `^` denotes negation, so `[^AZ]` matches any character except `A` or `Z`.

Your turn!
Find which countries' names start with a vowel and end in a consonant.

```{r}
# Fill in.
```
