---
title: "Lab 9"
author: "Your Name"
date: "July 14, 2015"
output: html_document
---

## Apply Functions

### Introduction

Apply functions are typically used to vectorize functions instead of using loops.
As you may know, many of the built-in functions in R are already vectorized.

```{r}
x <- 1:10
y <- sqrt(x)
print(y)
```

As you also know, this operation could have been performed with a simple loop.

```{r}
x <- 1:10
y <- numeric(length(x))
for (i in 1:length(x)) {
  y[i] <- sqrt(x[i])
}
print(y)
```

However, if the operation in question were not already vectorized, we could still avoid using a loop. A similar technique would work if `x` were a list instead of a vector.

```{r}
x <- 1:10
y <- sapply(x, sqrt)
print(y)
```

### lapply and sapply

The simplest apply function is lapply. lapply is typically used for simple one-argument functions that you want to apply to the elements of a vector or list. Built-in functions that are vectorized for vectors can also operate over lists with lapply.

```{r}
x <- as.list(1:10)
# y <- sqrt(x)    # What happens when you run this line?
y <- lapply(x, sqrt)
print(y)
```

sapply functions similarly to lapply but simplifies the output if possible.

```{r}
x <- as.list(1:10)
y <- sapply(x, sqrt)
print(y)
```

Of course, you can also use lapply and sapply with your own functions.

```{r}
sq <- function(x) {
  return(x^2)
}
x <- 1:10
y <- sapply(x, sq)
print(y)
```

Your turn!

The function below "squishes" a number into the interval [0, 100] (as in Lab 6).

```{r}
squish <- function(x) {
  if (x < 0) {
    x <- 0
  } else if (x > 100) {
    x <- 100
  }
  return(x)
}
```

Use an apply function to apply squish to an entire vector.

```{r}
x <- (1:12)^2 - 22
# Fill this in.
```

Here is a for loop which performs a certain operation.

```{r}
set.seed(0)
x <- rnorm(10)
y <- numeric(length(x))
for (i in 1:length(x)) {
  y[i] <- (x[i] - min(x)) / (max(x) - min(x))
}
```

Replace the above code with equivalent code which does not use a loop. You may wish to write an explicit or anonymous function.

```{r}
set.seed(0)
x <- rnorm(10)
# Fill this in.
```

### Other Basic Apply Functions

lapply and sapply are best suited for one-dimensional data structures (i.e. vectors and lists).
If you want to use an apply function for a multidimensional data structure, the simplest apply function is apply.
apply has an additional argument corresponding to the dimension (or dimensions) over which you want to apply the function.

```{r}
set.seed(0)
x <- matrix(rnorm(10), nrow = 2, ncol = 5)
y <- apply(x, 1, mean)
print(y)
z <- apply(x, 2, mean)
print(z)
```

tapply has an additional argument which can be one or more factors to apply separately.

```{r}
set.seed(0)
df <- data.frame(x = rnorm(10), ltr = factor(sample(c("a", "b", "c"), 10, replace = TRUE)), nmbr = sample(c(1, 2, 3), 10, replace = TRUE))
byltr <- tapply(df$x, df$ltr, mean)
print(byltr)
# If the second argument is not a factor, it will be coerced to a factor.
bynmbr <- tapply(df$x, df$nmbr, mean)
print(bynmbr)
# Multiple factors can be included in a list or data frame.
byboth <- tapply(df$x, df[, c("ltr", "nmbr")], mean)
print(byboth)
```

Your turn!

Perform the listed operations on the HairEyeColor dataset.

```{r, eval = FALSE}
# Remove eval = FALSE or set eval = TRUE when you have completed this code chunk.
hec <- HairEyeColor
print(hec)
# First create a two-dimensional version of the dataset called hec2d, which does not separate the data by sex.
hec2d <- # Fill this in.
# Now use an apply function to determine the most common eye color for each hair color.
# You may wish to write an explicit or anonymous function.
# Fill this in.
```

Finally, mapply is similar to sapply, but can use functions with multiple arguments.

```{r}
x <- 1:10
y <- 11:20
geommean <- function(x, y) {
  return(sqrt(x*y))
}
# This function will actually work without mapply, since sqrt and * are already vectorized.
z <- mapply(geommean, x, y)
print(z)
```

### Related Functions

aggregate works similarly to tapply on a greater variety of data structures.

```{r}
set.seed(0)
df <- data.frame(x = rnorm(10), y = rnorm(10), ltr = factor(sample(c("a", "b", "c"), 10, replace = TRUE)), nmbr = sample(c(1, 2, 3), 10, replace = TRUE))
z <- aggregate(df[, c("x", "y")], df[, c("ltr", "nmbr")], mean)
print(z)
```

sweep can be used to normalize datasets.

```{r}
x <- matrix(1:10, nrow = 2, ncol = 5)
xmean1 <- apply(x, 1, mean)
xmean2 <- apply(x, 2, mean)
# Subtract off the mean of each row/column.
y <- sweep(x, 1, xmean1)
print(y)
z <- sweep(x, 2, xmean2)
print(z)
# Multiply by the mean of each column (for some reason).
w <- sweep(x, 2, xmean1, FUN = "*")
print(w)
```
